Fig 4 and Supple figs

### 2020-05-26 analysis/figure-4.Rmd
progress on fig 4

---
title: "Fig5"
output: md_document
---

@@

source("../R/themes.R")
source("../R/functions.R")
#source("../R/wrangledata.R")

@@

@@

  df <- df %>% mutate(hiloPRL = fct_collapse(treatment, 
                                  lo = c("m.inc.d8", "m.inc.d3", "m.inc.d9", "inc.d3", "inc.d9", "lay", "bldg"),
                                  hi = c("inc.d17",   "m.inc.d17", "prolong" ,  "hatch" ,  "m.n2", "extend", "n5")),
                extint = fct_collapse(treatment, 
                                  eggs = c("inc.d3", "inc.d9", "inc.d17", "prolong", "lay"),
                                  chicks = c( "m.inc.d8", "hatch", "extend", "n5"),
                                  loss = c("m.inc.d3", "m.inc.d9", "m.n2", "m.inc.d17", "bldg")))
  df$extint <- factor(df$extint, levels = levelsextint)
  return(df)
}

@@

```{r vsd}
vsd_path <- "../results/DEseq2/manip/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = blk.s031.pu.d_female_gonad_prolong:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts") 
allvsd <- as_tibble(allvsd)

@@

# for plotting gene expression over time
getcandidatevsdmanip <- function(whichgenes, whichtissue, whichsex){
  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/manip/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
     dplyr::mutate(treatment = recode(treatment, "m.hatch" = "m.n2",
                                                  "extend.hatch" = "m.n2",
                                      "inc.prolong" = "prolong")) %>%

    dplyr::mutate(day = fct_collapse(treatment,
                                     "1" = c("inc.d3", "m.inc.d3" ),
                                     "2" = c("inc.d9", "m.inc.d9", "m.inc.d8"  ),
                                     "2" = c("inc.d17", "m.inc.d17", "prolong"  ),
                                     "4" = c("hatch", "m.n2", "extend"  )
                                     )) %>%
    dplyr::mutate(day = as.numeric(day)) %>%
    dplyr::mutate(manip = fct_collapse(treatment,
                                       "reference" = charlevels,
                                       "removal" = c("m.inc.d3", "m.inc.d9", "m.inc.d17","m.n2"),
                                       "replacement" = c("m.inc.d8", "prolong", "extend"))) %>%
    dplyr::mutate(manip = factor(manip, levels = c("reference", "removal", "replacement"))) %>%
    dplyr::select(sex, tissue, treatment, day,  manip, gene, samples, counts) %>%
    filter(tissue == whichtissue, sex %in% whichsex)  %>%
    drop_na()
  candidates$treatment <- factor(candidates$treatment, levels = alllevels)
  return(candidates)
}

hypvsd <- getcandidatevsdmanip(candidategenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsdmanip(candidategenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsdmanip(candidategenes, "gonad", sexlevels)
candidatevsd <- rbind(hypvsd, pitvsd)
candidatevsd <- rbind(candidatevsd, gonvsd)
head(candidatevsd)

hypvsd2 <- getcandidatevsdmanip(hypothesisDEGs, "hypothalamus", sexlevels)
pitvsd2 <- getcandidatevsdmanip(hypothesisDEGs, "pituitary", sexlevels)
gonvsd2 <- getcandidatevsdmanip(hypothesisDEGs, "gonad", sexlevels)
candidatevsd2 <- rbind(hypvsd2, pitvsd2)
candidatevsd2 <- rbind(candidatevsd2, gonvsd2)
head(candidatevsd2) 
```

@@

c1 <- plottsneelipsev2(hyptsne, hyptsne$treatment, allcolors) + 
  labs(subtitle = "Hypothalamus ", x = NULL)  
c2 <- plottsneelipsev2(hyptsne, hyptsne$extint, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
c3 <- plottsneelipsev2(hyptsne, hyptsne$hiloPRL, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) 
c4 <- plottsneelipsev2(pittsne, pittsne$treatment, allcolors ) + 
  labs(subtitle = "Pituitary ", x = NULL) 
c5 <- plottsneelipsev2(pittsne, pittsne$extint, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) 
c6 <- plottsneelipsev2(pittsne, pittsne$hiloPRL, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
c7 <- plottsneelipsev2(gontsne, gontsne$treatment, allcolors ) + 
  labs(subtitle = "Ggonads ")  
c8 <- plottsneelipsev2(gontsne, gontsne$extint, allcolors ) + 
  labs(subtitle = " " )    +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
c9 <- plottsneelipsev2(gontsne, gontsne$hiloPRL, allcolors ) + 
  labs(subtitle = " " ) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())

@@

candidatevsd <- candidatevsd %>%
  dplyr::mutate(treatment = fct_collapse(treatment,
                                         "m.n2" = c("m.n2",  "m.hatch" ),
                                         "prolong" = c("prolong", "inc.prolong") ,
                                         "extend" = c("extend", "extend.hatch")
                                         ))  %>%
  dplyr::mutate(day = fct_collapse(treatment,
                                    "0" = c("control", "bldg"),
                                    "1" = c("lay"),
                                     "3" = c("inc.d3", "m.inc.d3" ),
                                     "9" = c("inc.d9", "m.inc.d9", "m.inc.d8"  ),
                                     "17" = c("inc.d17", "m.inc.d17", "prolong"  ),
                                     "18" = c("hatch", "m.n2" ),
                                     "22" = c("n5",  "extend" ),
                                     "26" = c("n9"  )
                                     )) %>%
   mutate(treatment = factor(treatment, levels = alllevels)) %>%
  dplyr::mutate(day = factor(day, levels = c("0", "1", "3", "9", "17", "18", "22", "26")))

>>>>

---
title: "Fig 4 and Supple figs"
output: md_document
---

@@

source("../R/themes.R")
source("../R/functions.R")
source("../R/genelists.R")

@@

## import limma counts and sample info

```{r data}
countData <- read_csv("../results/01_limma.csv") %>%
  column_to_rownames(var = "X1")
countData <- as.data.frame(t(countData))
head(countData[1:3])

colData <- read_csv("../metadata/00_colData.csv") %>%
  mutate(treatment = factor(treatment, levels = alllevels),
         tissue = factor(tissue, levels = tissuelevels)) %>% 
  column_to_rownames(var = "X1") 
head(colData)
# check ready for analysis
# row.names(countData) == row.names(colData)
head(row.names(countData) == row.names(colData))
```

## tsne

@@

  df <- df %>% mutate(earlylate = fct_collapse(treatment, 
                                  early = c("early", "m.inc.d3", "m.inc.d9", "inc.d3", 
                                         "inc.d9", "lay", "bldg"),
                                  late = c("inc.d17",   "m.inc.d17", "prolong" ,  "hatch" , 
                                         "m.n2", "extend", "n5")),
                extint = fct_collapse(treatment, 
                                  eggs = c("inc.d3", "inc.d9", "inc.d17", "prolong", "lay"),
                                  chicks = c( "early", "hatch", "extend", "n5"),
                                  loss = c("m.inc.d3", "m.inc.d9", "m.n2", "m.inc.d17", "bldg")))
  df$extint <- factor(df$extint, levels = levelsextint)
  return(df)
}

@@

```{r vsd}
vsd_path <- "../results/DEseq2/hypothesis/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

## before pivoting, check names of df with
## head(names(allvsd)) and tail(names(allvsd))

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = blk.s061.pu.y_female_gonad_inc.d9:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts") 
allvsd <- as_tibble(allvsd)

@@

allvsd$samples
```

```{r candidatevsd}

# for plotting gene expression over time
getcandidatevsdmanip <- function(whichgenes, whichtissue){

  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1),
                  sextissue = sapply(strsplit(sextissue, '../results/DEseq2/hypothesis/'), "[", 2),
                  sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4),
                treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1),
                treatment = recode(treatment, "m.hatch" = "m.n2", "extend.hatch" = "m.n2",
                                                   "inc.prolong" = "prolong", "m.inc.d8" = "early"),
                treatment =  factor(candidates$treatment, levels = alllevels),
                  earlylate = fct_collapse(treatment, 
                                  early = c("early", "m.inc.d3", "m.inc.d9", "inc.d3", 
                                         "inc.d9", "lay", "bldg"),
                                  late = c("inc.d17",   "m.inc.d17", "prolong" ,  "hatch" , 
                                         "m.n2", "extend", "n5")),
                extint = fct_collapse(treatment, 
                                  eggs = c("inc.d3", "inc.d9", "inc.d17", "prolong", "lay"),
                                  chicks = c( "early", "hatch", "extend", "n5"),
                                  loss = c("m.inc.d3", "m.inc.d9", "m.n2", "m.inc.d17", "bldg")),
                manip = factor(manip, levels = c("reference", "removal", "replacement")))
  return(candidates)
} 
  %>%
    dplyr::select(gene, counts, sex, tissue, treatment, day, manip, earlylate, extint, samples)  %>%
    dplyr::filter(tissue == whichtissue) 
  return(candidates)

hypvsd <- getcandidatevsdmanip(c("PRL", "DRD1"), "hypothalamus")
pitvsd <- getcandidatevsdmanip(candidategenes, "pituitary")
gonvsd <- getcandidatevsdmanip(candidategenes, "gonad")
candidatevsd <- rbind(hypvsd, pitvsd)
candidatevsd <- rbind(candidatevsd, gonvsd)
head(candidatevsd)

```{r timingfigure, fig.height= 7.25in}

plotcandidatetiming <- function(df, whichgene, whichtissue){

  dfslim <- df %>%
    dplyr::filter(gene  == whichgene, treatment %in% timing) 
  print(head(dfslim))

  p <- dfslim %>%
    ggplot(aes(y =  counts, x = treatment, fill = treatment, color = sex)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size = 0.25, aes(color = sex)) +
    theme_B3() +
    theme(legend.position = "none",
          #axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          plot.subtitle = element_text(face = "italic")) +
    scale_color_manual(values = allcolors) +
    scale_fill_manual(values = allcolors)  +
    labs(title = paste(whichtissue, sep = " "), subtitle = whichgene, y = "gene expression") +
    facet_wrap(~sex)  +
    geom_signif(comparisons = list(c( "inc.d9", "early"),
                                   c( "inc.d17", "prolong"),
                                   c( "hatch", "prolong"),
                                   c( "hatch", "extend")),
                map_signif_level=TRUE,
                textsize = 3, family = 'Helvetica',
                vjust = 1.5, size = 0)
  return(p)

}

a <- plotcandidatetiming(hypvsd, "DRD1", "hypothalamus")
b <- plotcandidatetiming(pitvsd, "PRL", "pituitary")
c <- plotcandidatetiming(gonvsd, "CRHBP", "gonads")

plot_grid(a,b,c, nrow = 3)

@@

c1 <- plottsneelipsev2(hyptsne, hyptsne$treatment, allcolors) + 
  labs(subtitle = "Hypothalamus ", x = NULL)  
c2 <- plottsneelipsev2(hyptsne, hyptsne$extint, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
c3 <- plottsneelipsev2(hyptsne, hyptsne$earlylate, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) 
c4 <- plottsneelipsev2(pittsne, pittsne$treatment, allcolors ) + 
  labs(subtitle = "Pituitary ", x = NULL) 
c5 <- plottsneelipsev2(pittsne, pittsne$extint, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) 
c6 <- plottsneelipsev2(pittsne, pittsne$earlylate, allcolors ) + 
  labs(subtitle = " ", x = NULL) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
c7 <- plottsneelipsev2(gontsne, gontsne$treatment, allcolors ) + 
  labs(subtitle = "Ggonads ")  
c8 <- plottsneelipsev2(gontsne, gontsne$extint, allcolors ) + 
  labs(subtitle = " " )    +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
c9 <- plottsneelipsev2(gontsne, gontsne$earlylate, allcolors ) + 
  labs(subtitle = " " ) + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())

@@

candidatevsd <- candidatevsd %>%
  dplyr::mutate(treatment = fct_collapse(treatment,
                                         "m.n2" = c("m.n2",  "m.hatch" ),
                                         "prolong" = c("prolong", "inc.prolong") ,
                                         "extend" = c("extend", "extend.hatch")
                                         ))  %>%
  dplyr::mutate(day = fct_collapse(treatment,
                                    "0" = c("control", "bldg"),
                                    "1" = c("lay"),
                                     "3" = c("inc.d3", "m.inc.d3" ),
                                     "9" = c("inc.d9", "m.inc.d9", "early"  ),
                                     "17" = c("inc.d17", "m.inc.d17", "prolong"  ),
                                     "18" = c("hatch", "m.n2" ),
                                     "22" = c("n5",  "extend" ),
                                     "26" = c("n9"  )
                                     )) %>%
   mutate(treatment = factor(treatment, levels = alllevels)) %>%
  dplyr::mutate(day = factor(day, levels = c("0", "1", "3", "9", "17", "18", "22", "26")))
